---
title: "markers"
author: "Taran"
date: "2026-07-07"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


```{r} 
library(dplyr)
library(Seurat)
library(patchwork)
library(ggplot2)
library(DoubletFinder)
library(tidyverse)
library(RColorBrewer)
library(cowplot)
library(openxlsx)

# colour palette 
my_cols <- brewer.pal(n=5, "Spectral")


## feature and cell count function

FeatCellCount <- function(seuratObj){dim(seuratObj)
              }

#create a workbook that contains assigns each clusters data to its own sheet

cluster_workbook <- function(data, n_clusters){
  
  wb <- createWorkbook()
  addWorksheet(wb, "Markers")
writeData(wb, sheet = "Markers", x = filtered_markers)

  for(i in 0:n_clusters){
    
    cluster_data <- filtered_markers[filtered_markers$cluster == i,] 
    addWorksheet(wb, as.character(i))
    writeData(wb, sheet = as.character(i), x = cluster_data)
    saveWorkbook(wb, file = "filtered_markers.xlsx", overwrite = TRUE)
    
  }
  
  
}



```

# find markers for every cluster compared to all remaining cells, report only the positive ones, write to excel file
```{r}
markers <- FindAllMarkers(HH28A, only.pos = TRUE, group.by = "RNA_snn_res.0.6")
markers %>%
    group_by(cluster) %>%
    dplyr::filter(avg_log2FC > 1)


write.xlsx(markers, file = "/nobackup/proj/comet_icnlmdhuc/Taran/chicken_RNA_seq/HH28/sample_1/reg-diff-markers/HH28A-reg-diff-markers.xlsx")

```


#filter out "ENSGAL" genes

```{r}

filtered_markers <- markers %>% filter(!startsWith(gene, "ENSGAL"))

```



#write the markers to a workbook

```{r}

cluster_workbook(filtered_markers, 12)

```



#several tests for differential expression - set parameter test.use = "ROC" bc it will return the classification power for any individual marker
#find differential expression of markers for cluster 0 and write to excel file

```{r}
cluster0.markers <- FindMarkers(HH28A, ident.1 = 0, logfc.threshold = 0.25, test.use = "roc", only.pos = TRUE)

write.xlsx(cluster0.markers, file = "HH28A_c0_markers.xlsx")
```


#look for marker expression throughout UMAP plot to identify clusters

```{r}

Progenitor_markers <- FeaturePlot(HH28A, features = c("ATOH7", "HES6"))

amacrine <- FeaturePlot(HH28A, features = c("TFAP2A", "TFAP2B", "PROX1", "LHX1"))

early_cone_photoreceptors <- FeaturePlot(HH28A, features = c("RAX2", "RXRG", "CRX"))

neural_retina <- FeaturePlot(HH28A, features = c("SMOC1", "ALDH1A3", "VAX1", "VSX2", "CDH2"))

related_to_RPE_and_RPC <- FeaturePlot(HH28A, features = c("TYR", "TYRP", "WNT2B", "BCN"))

retinal_ganglion_cells <- FeaturePlot(HH28A, features = c("ELAVL4", "NEFL", "NEFM", "CL12", "GAP43", "STMN", "POU4F3", "SATB1", "SATB2", "PAX6", "RBPMS2"))

retinal_progenitor_cells <- FeaturePlot(HH28A, features = c("ZIC1", "VSX2", "PAX6", "VAX1", "SMOC1", "PMPR1B", "ATOH7", "HES6"))

endothelial_and_blood <- FeaturePlot(HH28A, features = c("VWF", "KDR"))

POM <- FeaturePlot(HH28A, features = c("PITX2", "EMX2", "LMX1B", "PAX7", "ALX1", "LMX1B", "ALX4", "DCN", "LTBP1", "COL3A1"))

RPE <- FeaturePlot(HH28A, features = c("OTX2", "MITF", "TYR", "TYRP1", "PITX2", "PMEL", "CDH1"))

pioneer <- FeaturePlot(HH28A, features = c("PAX2", "CLYBL", "NTN1", "MET", "EGF", "DVL1", "GALNT6", "FOXP2", "CDH2"))

blood <- FeaturePlot(HH28A, features = c("HBE1", "HBBA", "HBAD", "HBA1", "HBE", "HBZ", "HBBR", "SLC4A1", "SLC39A8", "TSPO2"))


cmz <- FeaturePlot(HH28A, features = c("COL9A1", "CPAMD8", "MECOM", "OPTC", "RELN", "BNC", "FBN2", "FBN3"))




```

```{r}

Progenitor_markers 

amacrine 

early_cone_photoreceptors 

neural_retina 

related_to_RPE_and_RPC 

retinal_ganglion_cells 

retinal_progenitor_cells 

endothelial_and_blood 

POM 

RPE 

pioneer 

blood


cmz 


```

#assign cluster identity


```{r}
new.cluster.ids <- c("NR", "RPE", "RPE2", "NR2", "NR3", "NR4", "CMZ", "NR5", "Pioneer", "RPC", "POM", "EC", "Blood")


Idents(HH28A) <- HH28A@meta.data$RNA_snn_res.0.6

names(new.cluster.ids) <- levels(HH28A)

HH28A <- RenameIdents(HH28A, new.cluster.ids)

```



```{r}
DimPlot(HH28A, reduction = "umap", label = TRUE, label.size = 10) 
ggsave("/nobackup/proj/comet_icnlmdhuc/Taran/chicken_RNA_seq/HH28/sample_1/plots/cc/reg-diff/ant_reg_diff.png", dpi = 400, width =45, height = 25, units = "cm")


```



```{r}

saveRDS(HH28A, file = "/nobackup/proj/comet_icnlmdhuc/Taran/chicken_RNA_seq/HH28/sample_1/sObj/sObj_8_full_reg_ant.RDS")

```



```{r}

saveRDS(HH28A, file = "/nobackup/proj/comet_icnlmdhuc/Taran/chicken_RNA_seq/HH28/sample_1/sObj/sObj_9_diff_reg_ant.RDS")

```
