---
title: "setup seurat object"
output: html_document
date: "2025-11-25"
fig_width: 6
fig_height: 4
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

#load all libaries required for this script

```{r} 
library(dplyr)
library(Seurat)
library(patchwork)
library(ggplot2)
library(DoubletFinder)
library(tidyverse)
library(RColorBrewer)
library(cowplot)
library(openxlsx)

# colour palette 
my_cols <- brewer.pal(n=5, "Spectral")


## feature and cell count function

FeatCellCount <- function(seuratObj){dim(seuratObj)
              }
  
  
##create function for doublet finder 


preDoublets <- function(seuratObj){  
    seuratObj <- NormalizeData(seuratObj)
    
    seuratObj <- ScaleData(seuratObj)
    seuratObj <- FindVariableFeatures(seuratObj, 
                                      x.low.cutoff = 0.0125, 
                                      y.cutoff = 0.25, 
                                      do.plot=FALSE)
    seuratObj <- RunPCA(seuratObj, pc.genes = seuratObj@var.genes, pcs.print = 0)
    
 
    return(seuratObj)
}



findDoublets <- function(seuratObj, n_pcs=30){
    
   
    nExp_poi  <- round(0.15*nrow(seuratObj@meta.data))
    
    print('yay!')
  
    seuratObj <- doubletFinder(seuratObj, 
                                  PCs = 1:30, 
                                  pN = 0.25, 
                                  pK = 0.01, 
                                  nExp = nExp_poi, 
                                  sct=FALSE)
    
    
    
     seuratObj@meta.data <- seuratObj@meta.data %>% 
        rename_at(vars(starts_with("DF.classifications")), 
                  funs(str_replace(., ".*", "DF.classifications"))) %>%
        rename_at(vars(starts_with("pANN")), 
                  funs(str_replace(., ".*", "pANN")))
    
       
  
        return(seuratObj) 
    
}


#create a workbook that contains assigns each clusters data to its own sheet

cluster_workbook <- function(data, n_clusters){
  
  wb <- createWorkbook()
  addWorksheet(wb, "Markers")
writeData(wb, sheet = "Markers", x = filtered_markers)

  for(i in 0:n_clusters){
    
    cluster_data <- filtered_markers[filtered_markers$cluster == i,] 
    addWorksheet(wb, as.character(i))
    writeData(wb, sheet = as.character(i), x = cluster_data)
    saveWorkbook(wb, file = "filtered_markers.xlsx", overwrite = TRUE)
    
  }
  
  
}


#top 5 genes for each cluster, sorted largest to smallest for avg log, more than 0.1 for pct.1 and less than 0.05 for p val adj
top5 <- function(sObj){

clusters <- filtered_markers %>% group_by(cluster) %>% arrange(cluster, desc(avg_log2FC)) %>% filter(pct.1 > 0.1) %>% filter(p_val_adj < 0.05) %>% top_n(5, avg_log2FC)
  

return(clusters)  
 
   
}

#compare umap plots for 5 resolutions, select the one that provides an accurate representation of the biological data

umap_trial <- function(sObj, resolutions = c(0.6, 0.7, 0.8, 0.9, 1),
                           dims, label.size = 10, repel = TRUE){
  
  plots <- list()
  objects <- list()
  
  for(res in resolutions){
    
    
    objres <- FindClusters(sObj, resolution = res)
    objres <- RunUMAP(objres, dims = dims)
    
    
    
    plots[[as.character(res)]] <- DimPlot(
      objres, reduction = "umap", label = TRUE, label.size = 10, 
      repel = TRUE) + ggtitle(paste0("Resolution = ", res))
    
    objects[[as.character(res)]] <- objres
  }
  return(list(plots = plots, objects = objects))
}




```

#load scRNA-seq data set 

```{r}

HH28D.data <- Read10X(data.dir = "/nobackup/proj/comet_icnlmdhuc/Taran/chicken_RNA_seq/results/Sample_4_Results/outs/filtered_feature_bc_matrix/")
```

#assign to seurat object

```{r}

HH28D <- CreateSeuratObject(counts = HH28D.data, project = "HH28", min.cells = 3, min.features = 200)
HH28D


```


#add column to metadata to assign the percentage of mitochondrial dna to

```{r}
HH28D[["percent.mt"]] <- PercentageFeatureSet(HH28D, pattern = "^MT-")
```





#visualise QC metrics as violin plot 

```{r} eval=FALSE
#only for paper, rest is better for accurate subsetting
QC_metric<- VlnPlot(HH28D, features = c("nFeature_RNA", "nCount_RNA", "percent.mt"), ncol = 3)
QC_metric

```


#feature
```{r}
QC_mf<- VlnPlot(HH28D, features = "nFeature_RNA")
QC_mf
```

```{r}
QC_metric_feature<- VlnPlot(HH28D, features = "nFeature_RNA", y.max = 1000)
QC_metric_feature
```

#counts
```{r}
QC_mc<- VlnPlot(HH28D, features = "nCount_RNA")
QC_mc
```

```{r}
QC_metric_counts<- VlnPlot(HH28D, features = "nCount_RNA", y.max = 1500)
QC_metric_counts
```

#mitochondrial RNA
```{r}
QC_mm<- VlnPlot(HH28D, features = "percent.mt")
QC_mm
```

```{r}
QC_metric_mito<- VlnPlot(HH28D, features = "percent.mt", y.max = 10)
QC_metric_mito
```



#feature scatter used to visualise the feature-feature relationships but can be used for any relationship

```{r}

mt<- FeatureScatter(HH28D, feature1 = "nCount_RNA", feature2 = "percent.mt")
features <- FeatureScatter(HH28D, feature1 = "nCount_RNA", feature2 = "nFeature_RNA")
mtfeat <- FeatureScatter(HH28D, feature1 = "nFeature_RNA", feature2 = "percent.mt")

mt+ features + mtfeat
```






#subset data to look at specific parameters bc values that are out of scope aren't important and therefore not needed in the graph (??? help)
#remove mitochondrial genes


```{r}
HH28D <- subset(HH28D, subset = nFeature_RNA > 800 & 
                                percent.mt   <5     &
                                nCount_RNA   >1300  
                )

mt2 <- FeatureScatter(HH28D, feature1 = "nCount_RNA", feature2 = "percent.mt")
features2 <- FeatureScatter(HH28D, feature1 = "nCount_RNA", feature2 = "nFeature_RNA")

mt2 + features2

feat2 <- VlnPlot(HH28D, features = "nFeature_RNA", y.max = 4000)

feat2

count <- VlnPlot(HH28D, features = "nCount_RNA", y.max = 10000)

count
```
#compare mitchondrial RNA subset data

```{r}h
mt_genes_scatter_plot1 + mt_genes_scatter_plot3
```

#compare features subset data

```{r}h
features_scatter_plot2 + features_scatter_plot4

```

#nuclei after subset

```{r}
HH28D

```


#doubletFinder

```{r}

HH28D <- preDoublets(HH28D)
HH28D <- findDoublets(HH28D)

```

```{r}
doublet <- FeatureScatter(HH28D, feature1 = "nCount_RNA", feature2 = "nFeature_RNA", group.by = "DF.classifications")

doublet
```

```{r}
HH28D <- subset(HH28D, subset = DF.classifications != "Doublet" )
```



#plot after doublet finder comparison

```{r}h
mt_genes_scatter_plot5 <- FeatureScatter(HH28D, feature1 = "nCount_RNA", feature2 = "percent.mt")
features_scatter_plot6 <- FeatureScatter(HH28D, feature1 = "nCount_RNA", feature2 = "nFeature_RNA")

```

#compare mito RNA before and after subset and after doublet finder

```{r}h

mt_genes_scatter_plot1 + mt_genes_scatter_plot3 + mt_genes_scatter_plot5

```

```{r}h

features_scatter_plot2 + features_scatter_plot4 + features_scatter_plot6
```

#nuclei after removal

```{r}
HH28D
```



#checkpoint after QC 

```{r}
saveRDS(HH28D, file = "sObj_QC_1.RDS")
```




#normalise the data

```{r}
HH28D <- NormalizeData(HH28D, normalization.method = "LogNormalize", scale.factor = 10000)
```

#find the variable features


```{r}
HH28D <- FindVariableFeatures(HH28D, selection.method = "vst", nfeatures = 2000)
```

#identify top 10 variable features

```{r}
top10 <- head(VariableFeatures(HH28D), 10)
```

#plot variable features

```{r}

variable_feat_plot <- VariableFeaturePlot(HH28D)
labelled_variable_feat_plot <- LabelPoints(plot = variable_feat_plot, points = top10, repel = TRUE, xnudge = 0, ynudge = 0)

```

#view plot

```{r}
variable_feat_plot +labelled_variable_feat_plot
```

# scale data so values are workable and are of same meaning

```{r}
all.genes <- rownames(HH28D)
HH28D <- ScaleData(HH28D, features = all.genes)
```


#run a principle component analysis to get our principle components

```{r}
HH28D <- RunPCA(HH28D, features = VariableFeatures(object = HH28D))
```

#look at first 5 PCAs

```{r}
print(HH28D[["pca"]], dims = 1:5, nFeatures = 5)
```

#do a heatmap to look at first PC

```{r}

DimHeatmap(HH28D, dims = 1, cells = 500, balanced = TRUE, fast = FALSE) + scale_fill_gradientn(colours = my_cols)



```


```{r}

PC1t50 <- DimHeatmap( HH28D, dims = 1:10,cells = 500, balanced  = TRUE, fast = FALSE, combine = FALSE)

PC1t5_col <- lapply(PC1t5, function(p) p + scale_fill_gradientn(colours = my_cols)) 

wrap_plots(plotlist = PC1t5_col, nrow = 3, ncol = 2)


```

#checkpoint after PCA

```{r}

saveRDS(HH28D, file = "sObj_PCA_2.RDS")

```


#elbow plot to determine dimensionality for the cluster map

```{r}
Elbow_plot <- ElbowPlot(HH28D, ndims = 40)
Elbow_plot
```



# find the neighbours 

```{r}
HH28D <- FindNeighbors(HH28D, dims = 2:30)

```

#find clusters
```{r}
HH28D <- FindClusters(HH28D, resolution = c(0.6, 0.7, 0.8, 0.9))

```



#create the UMAP data - this the clusters the cells with common gene expression and links it to the neighbours

```{r}

HH28D <- RunUMAP(HH28D, dims = 2:30)
```

#visualise the UMAP

```{r}

res <- "RNA_snn_res."
nums = c(0.6, 0.7, 0.8, 0.9)
paste0(res, nums)

options(bitmapType = "cairo")
png("umap_trial.png", width = 2400, height = 2000, res = 200, type = "cairo")
DimPlot(HH28D, reduction = "umap", group.by = paste0(res, nums), label = TRUE, label.size = 6, repel = TRUE)
dev.off()



```

#save as png
```{r}



png(file= "umap_HH28D.png", width=2000, height=1500, res = 200, type = "cairo") 
DimPlot(HH28D, reduction = "umap", group.by = "RNA_snn_res.0.6", label = TRUE, label.size = 5, repel = TRUE) 
dev.off()

```

#save RDS

```{r}
saveRDS(HH28D, file = "sObj_UMAP_3.RDS")
```



# find markers for every cluster compared to all remaining cells, report only the positive ones, write to excel file
```{r}
markers <- FindAllMarkers(HH28D, only.pos = TRUE, group.by = "RNA_snn_res.0.6")
markers %>%
    group_by(cluster) %>%
    dplyr::filter(avg_log2FC > 1)


write.xlsx(markers, file = "HH28D_markers.xlsx")

```


#filter out "ENSGAL" genes

```{r}

filtered_markers <- markers %>% filter(!startsWith(gene, "ENSGAL"))

```



#write the markers to a workbook

```{r}

cluster_workbook(filtered_markers, 13)

```



#several tests for differential expression - set parameter test.use = "ROC" bc it will return the classification power for any individual marker
#find differential expression of markers for cluster 0 and write to excel file

```{r}
cluster0.markers <- FindMarkers(HH28D, ident.1 = 0, logfc.threshold = 0.25, test.use = "roc", only.pos = TRUE)

write.xlsx(cluster0.markers, file = "HH28D_c0_markers.xlsx")
```



#look for marker expression throughout UMAP plot to identify clusters

```{r}

Progenitor_markers <- FeaturePlot(HH28D, features = c("ATOH7", "HES6"))

amacrine <- FeaturePlot(HH28D, features = c("TFAP2A", "TFAP2B", "PROX1", "LHX1"))

early_cone_photoreceptors <- FeaturePlot(HH28D, features = c("RAX2", "RXRG", "CRX"))

neural_retina <- FeaturePlot(HH28D, features = c("SMOC1", "ALDH1A3", "VAX1", "VSX2", "CDH2"))

related_to_RPE_and_RPC <- FeaturePlot(HH28D, features = c("TYR", "TYRP", "WNT2B", "BCN"))

retinal_ganglion_cells <- FeaturePlot(HH28D, features = c("ELAVL4", "NEFL", "NEFM", "CL12", "GAP43", "STMN", "POU4F3", "SATB1", "SATB2", "PAX6", "RBPMS2"))

retinal_progenitor_cells <- FeaturePlot(HH28D, features = c("ZIC1", "VSX2", "PAX6", "VAX1", "SMOC1", "PMPR1B", "ATOH7", "HES6"))

endothelial_and_blood <- FeaturePlot(HH28D, features = c("VWF", "KDR"))

POM <- FeaturePlot(HH28D, features = c("PITX2", "EMX2", "LMX1B", "PAX7", "ALX1", "LMX1B", "ALX4", "DCN", "LTBP1", "COL3A1"))

RPE <- FeaturePlot(HH28D, features = c("OTX2", "MITF", "TYR", "TYRP1", "PITX2", "PMEL", "CDH1"))

pioneer <- FeaturePlot(HH28D, features = c("PAX2", "CLYBL", "NTN1", "MET", "EGF", "DVL1", "GALNT6", "FOXP2", "CDH2"))

blood <- FeaturePlot(HH28D, features = c("HBE1", "HBBA", "HBAD", "HBA1", "HBE", "HBZ", "HBBR", "SLC4A1", "SLC39A8", "TSPO2"))


cmz <- FeaturePlot(HH28D, features = c("COL9A1", "CPAMD8", "MECOM", "OPTC", "RELN", "BNC", "FBN2", "FBN3"))




```

```{r}

Progenitor_markers 

amacrine 

early_cone_photoreceptors 

neural_retina 

related_to_RPE_and_RPC 

retinal_ganglion_cells 

retinal_progenitor_cells 

endothelial_and_blood 

POM 

RPE 

pioneer 

blood


cmz 


```

#top 5 genes for each cluster, sorted largest to smalled avg log and less than 0.05 for p val adj
```{r}
top5.genes <- top5(HH28D)

top5.genes
``` 


#heatmap for the top 5 genes per cluster

```{r}
top5heatmap <- DoHeatmap(HH28D, group.by = "RNA_snn_res.0.6", features = top5.genes$gene)
top5heatmap

```


```{r}

options(bitmapType = "cairo")
png("top5genes.png", width = 9000, height = 4000, res = 200, type = "cairo") 
p <- DoHeatmap(HH28D, group.by = "RNA_snn_res.0.7", features = top5.genes$gene, label = FALSE) 
p <- p + theme(
  axis.text = element_text(size = 20),         #gene names
  legend.title = element_text(size = 20),      #identity and expression text size
  legend.text = element_text(size = 20),       #key label size
  legend.key.size = unit(1.5, "cm"))           #box size
  
wrap_plots(p)

dev.off()
```



#assign cluster identity


```{r}
new.cluster.ids <- c("0", "1", "2", "RPE", "CMZ", "5", "RPE2", "Pioneer", "8","POM","RPC", "EC", "12", "13" )


Idents(HH28D) <- HH28D@meta.data$RNA_snn_res.0.6

names(new.cluster.ids) <- levels(HH28D)

HH28D <- RenameIdents(HH28D, new.cluster.ids)
DimPlot(HH28D, reduction = "umap", label = TRUE, label.size = 8) + xlab("UMAP 1") + ylab("UMAP 2") +
    theme(axis.title = element_text(size = 18), legend.text = element_text(size = 18)) + guides(colour = guide_legend(override.aes = list(size = 10)))
```




```{r}

png(file= "umap_HH28D.png", width=2800, height=1500, res = 200, type = "cairo") 
DimPlot(HH28D, reduction = "umap", label = TRUE, label.size = 8) + xlab("UMAP 1") + ylab("UMAP 2") +
    theme(axis.title = element_text(size = 18), legend.text = element_text(size = 18)) + guides(colour = guide_legend(override.aes = list(size = 10)))
dev.off()


```



```{r}

saveRDS(HH28D, file = "sObj_Markers_4.RDS")

```




#segregate list of cell cycle markers into G2/M and S

```{r}
s.genes <- cc.genes$s.genes
g2m.genes <- cc.genes$g2m.genes
```

#assign cell-cycle scores

```{r}
HH28D <- CellCycleScoring(HH28D, s.features = s.genes, g2m.features = g2m.genes, set.ident = TRUE)

```

#visualise distribution
```{r}

png("cell_cycle_distribution.png", width = 2600, height = 2000)

plots <- RidgePlot(HH28D, 
                   features = c("PCNA","TOP2A","MCM6","MKI67"), ncol = 2, 
                   cols = c("S" = "orange", "G1" = "forestgreen", "G2M" = "steelblue4"), 
                   combine = FALSE)

plots <- lapply(plots, function(p) p + 
                  theme(plot.title = element_text(size = 40),
                        axis.text = element_text(size = 30),
                        axis.title.x = element_text(size = 30),
                        axis.title.y = element_text(size = 30),
                        legend.text = element_text(size = 30)) + 
                  guides(colour = guide_legend(override.aes = list(size = 30))))
                
wrap_plots(plots, ncol = 2)   

dev.off()

```

#update how clusters are ordered in metadata to ensure uniform ordering across all sample reports
```{r}
# Update the identity of the Seurat object and define the new order of groups
HH28D@meta.data$Phase <- factor(HH28D@meta.data$Phase, levels = c("G1", "S", "G2M"))


# Reassign identity with the new order
Idents(HH28D) <- HH28D@meta.data$Phase
```



```{r}

png("cell_cycle_plot.png", width = 1600, height = 1000)
DimPlot(HH28D, cols = c("S" = "orange", "G1" = "forestgreen", "G2M" = "steelblue4")) +
    theme(axis.title = element_text(size = 18), legend.text = element_text(size = 18)) + guides(colour = guide_legend(override.aes = list(size = 10)))
dev.off()

```


```{r}
saveRDS(HH28D, file = "sObj_cell_cycle_5.RDS")
```


