---
title: "cell-cycle-regression"
author: "Taran"
date: "2026-07-07"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


```{r} 
library(dplyr)
library(Seurat)
library(patchwork)
library(ggplot2)
library(DoubletFinder)
library(tidyverse)
library(RColorBrewer)
library(cowplot)
library(openxlsx)

# colour palette 
my_cols <- brewer.pal(n=5, "Spectral")


## feature and cell count function

FeatCellCount <- function(seuratObj){dim(seuratObj)
              }


```


#segregate list of cell cycle markers into G2/M and S

```{r}
s.genes <- cc.genes$s.genes
g2m.genes <- cc.genes$g2m.genes
```


#############################################################################   regress signal completely

```{r}

HH28D <- ScaleData(HH28D, vars.to.regress = c("S.Score", "G2M.Score"), features = VariableFeatures(HH28D)) 

```


#PCA on the variable genes no longer returns components associated with cell cycle

```{r}
HH28D <- RunPCA(HH28D, features = VariableFeatures(HH28D), nfeatures.print = 10)
```


```{r}
Elbow_plot <- ElbowPlot(HH28D, ndims = 40)
Elbow_plot
```


# find the neighbours 

```{r}
HH28D <- FindNeighbors(HH28D, dims = 1:30)

```

#find clusters
```{r}
HH28D <- FindClusters(HH28D, resolution = 0.7)

```


#create the UMAP data - this the clusters the cells with common gene expression and links it to the neighbours

```{r}

HH28D <- RunUMAP(HH28D, dims = 1:30)
```


#update how clusters are ordered in metadata to ensure uniform ordering across all sample reports
```{r}
# Update the identity of the Seurat object and define the new order of groups
HH28D@meta.data$Phase <- factor(HH28D@meta.data$Phase, levels = c("G1", "S", "G2M"))


# Reassign identity with the new order
Idents(HH28D) <- HH28D@meta.data$Phase
```

#plot

```{r}

DimPlot(HH28D, cols = c("S" = "orange", "G1" = "forestgreen", "G2M" = "steelblue4", group_by = "Phase")) 
ggsave("/nobackup/proj/comet_icnlmdhuc/Taran/chicken_RNA_seq/HH28/sample_4/plots/cc_reg.png", dpi = 400)

```


```{r}
saveRDS(HH28D, file = "/nobackup/proj/comet_icnlmdhuc/Taran/chicken_RNA_seq/HH28/sample_4/sObj/sObj_6_full_reg.RDS")
```


############################################################################################ regress difference


#find difference between scores
```{r}
HH28D$CC.Difference <- HH28D$S.Score - HH28D$G2M.Score
```

#regress the difference
```{r}
HH28D <- ScaleData(HH28D, vars.to.regress = "CC.Difference", features = VariableFeatures(HH28D))
```

# cell cycle effects removed in PCA
```{r}
HH28D <- RunPCA(HH28D, features = VariableFeatures(HH28D), nfeatures.print = 10)
```



```{r}
Elbow_plot <- ElbowPlot(HH28D, ndims = 40)
Elbow_plot
```


# find the neighbours 

```{r}
HH28D <- FindNeighbors(HH28D, dims = 1:30)

```

#find clusters
```{r}
HH28D <- FindClusters(HH28D, resolution = 0.7)

```

#create the UMAP data - this clusters the cells with common gene expression and links it to the neighbours

```{r}

HH28D <- RunUMAP(HH28D, dims = 1:30)
```

#update how clusters are ordered in metadata to ensure uniform ordering across all sample reports
```{r}
# Update the identity of the Seurat object and define the new order of groups
HH28D@meta.data$Phase <- factor(HH28D@meta.data$Phase, levels = c("G1", "S", "G2M"))


# Reassign identity with the new order
Idents(HH28D) <- HH28D@meta.data$Phase
```


#plot

```{r}

DimPlot(HH28D, cols = c("S" = "orange", "G1" = "forestgreen", "G2M" = "steelblue4", group_by = "Phase")) 
ggsave("/nobackup/proj/comet_icnlmdhuc/Taran/chicken_RNA_seq/HH28/sample_4/plots/cc_reg_diff.png", dpi = 400)


```


```{r}
saveRDS(HH28D, "/nobackup/proj/comet_icnlmdhuc/Taran/chicken_RNA_seq/HH28/sample_4/sObj/sObj_7_reg_diff.RDS")
```

