---
title: "setup seurat object"
output: html_document
date: "2025-11-25"
fig_width: 6
fig_height: 4
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

#load all libaries required for this script

```{r} 
library(dplyr)
library(Seurat)
library(patchwork)
library(ggplot2)
library(DoubletFinder)
library(tidyverse)
library(RColorBrewer)
library(cowplot)
library(openxlsx)

# colour palette 
my_cols <- brewer.pal(n=5, "Spectral")


## feature and cell count function

FeatCellCount <- function(seuratObj){dim(seuratObj)
              }
  
  
##create function for doublet finder 


preDoublets <- function(seuratObj){  
    seuratObj <- NormalizeData(seuratObj)
    
    seuratObj <- ScaleData(seuratObj)
    seuratObj <- FindVariableFeatures(seuratObj, 
                                      x.low.cutoff = 0.0125, 
                                      y.cutoff = 0.25, 
                                      do.plot=FALSE)
    seuratObj <- RunPCA(seuratObj, pc.genes = seuratObj@var.genes, pcs.print = 0)
    
 
    return(seuratObj)
}



findDoublets <- function(seuratObj, n_pcs=30){
    
   
    nExp_poi  <- round(0.15*nrow(seuratObj@meta.data))
    
    print('yay!')
  
    seuratObj <- doubletFinder(seuratObj, 
                                  PCs = 1:30, 
                                  pN = 0.25, 
                                  pK = 0.01, 
                                  nExp = nExp_poi, 
                                  sct=FALSE)
    
    
    
     seuratObj@meta.data <- seuratObj@meta.data %>% 
        rename_at(vars(starts_with("DF.classifications")), 
                  funs(str_replace(., ".*", "DF.classifications"))) %>%
        rename_at(vars(starts_with("pANN")), 
                  funs(str_replace(., ".*", "pANN")))
    
       
  
        return(seuratObj) 
    
}


```


#load scRNA-seq data set 

```{r}

HH28D.data <- Read10X(data.dir = "/nobackup/proj/comet_icnlmdhuc/Taran/chicken_RNA_seq/results/Sample_4_Results/outs/filtered_feature_bc_matrix/")
```

#assign to seurat object

```{r}

HH28D <- CreateSeuratObject(counts = HH28D.data, project = "HH28", min.cells = 3, min.features = 200)
HH28D


```


#add column to metadata to assign the percentage of mitochondrial dna to

```{r}
HH28D[["percent.mt"]] <- PercentageFeatureSet(HH28D, pattern = "^MT-")
```

#visualise QC metrics as violin plot 

```{r} 
#only for paper, rest is better for accurate sub-setting
QC_metric<- VlnPlot(HH28D, features = c("nFeature_RNA", "nCount_RNA", "percent.mt"), ncol = 3)
QC_metric

```


#feature
```{r}
QC_mf<- VlnPlot(HH28D, features = "nFeature_RNA")
QC_mf
```

```{r}
QC_metric_feature<- VlnPlot(HH28D, features = "nFeature_RNA", y.max = 1000)
QC_metric_feature
```

#counts
```{r}
QC_mc<- VlnPlot(HH28D, features = "nCount_RNA")
QC_mc
```

```{r}
QC_metric_counts<- VlnPlot(HH28D, features = "nCount_RNA", y.max = 1500)
QC_metric_counts
```

#mitochondrial RNA
```{r}
QC_mm<- VlnPlot(HH28D, features = "percent.mt")
QC_mm
```

```{r}
QC_metric_mito<- VlnPlot(HH28D, features = "percent.mt", y.max = 10)
QC_metric_mito
```



#feature scatter used to visualise the feature-feature relationships but can be used for any relationship

```{r}

mt<- FeatureScatter(HH28D, feature1 = "nCount_RNA", feature2 = "percent.mt")
features <- FeatureScatter(HH28D, feature1 = "nCount_RNA", feature2 = "nFeature_RNA")
mtfeat <- FeatureScatter(HH28D, feature1 = "nFeature_RNA", feature2 = "percent.mt")

mt+ features

mtfeat
```






#subset data to look at specific parameters bc values that are out of scope aren't important and therefore not needed in the graph (??? help)
#remove mitochondrial genes

```{r}
HH28D <- subset(HH28D, subset = nFeature_RNA > 800 & 
                                percent.mt   <5     &
                                nCount_RNA   >1300  
                )

mt2 <- FeatureScatter(HH28D, feature1 = "nCount_RNA", feature2 = "percent.mt")
features2 <- FeatureScatter(HH28D, feature1 = "nCount_RNA", feature2 = "nFeature_RNA")

mt2 + features2

feat2 <- VlnPlot(HH28D, features = "nFeature_RNA", y.max = 4000)

feat2

count <- VlnPlot(HH28D, features = "nCount_RNA", y.max = 10000)

count
```

#compare mitchondrial RNA subset data

```{r}
mt + mt2
```

#compare features subset data

```{r}
feat + feat2

```


#nuclei after subset

```{r}
HH28D
FeatCellCount(HH28D)
```

#doubletFinder

```{r}

HH28D <- preDoublets(HH28D)
HH28D <- findDoublets(HH28D)

```

```{r}
doublet <- FeatureScatter(HH28D, feature1 = "nCount_RNA", feature2 = "nFeature_RNA", group.by = "DF.classifications")

doublet
```

```{r}
HH28D <- subset(HH28D, subset = DF.classifications != "Doublet" )
```


```{r}
 doubsub<- FeatureScatter(HH28D, feature1 = "nCount_RNA", feature2 = "nFeature_RNA", group.by = "DF.classifications")
doubsub
```

#plot after doublet finder comparison

```{r}
mt3<- FeatureScatter(HH28D, feature1 = "nCount_RNA", feature2 = "percent.mt")
feat3 <- FeatureScatter(HH28D, feature1 = "nCount_RNA", feature2 = "nFeature_RNA")

```

#compare mito RNA before and after subset and after doublet finder

```{r}

mt + mt2 + mt3

```

```{r}

feat + feat2 + feat3
```


#nuclei after removal

```{r}
HH28D
```


#checkpoint after QC 

```{r}
saveRDS(HH28D, file = "sObj_QC_1.RDS")
```
