---
title: "reg-full"
author: "Taran"
date: "2026-08-04"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```



```{r} 
library(dplyr)
library(Seurat)
library(patchwork)
library(ggplot2)
library(DoubletFinder)
library(tidyverse)
library(RColorBrewer)
library(cowplot)
library(openxlsx)

# colour palette 
my_cols <- brewer.pal(n=5, "Spectral")


## feature and cell count function

FeatCellCount <- function(seuratObj){dim(seuratObj)
              }


```


#segregate list of cell cycle markers into G2/M and S

```{r}
s.genes <- cc.genes$s.genes
g2m.genes <- cc.genes$g2m.genes
```



```{r}

HH30F <- ScaleData(HH30F, vars.to.regress = c("S.Score", "G2M.Score"), features = VariableFeatures(HH30F)) 

```


#PCA on the variable genes no longer returns components associated with cell cycle

```{r}
HH30F <- RunPCA(HH30F, features = VariableFeatures(HH30F), nfeatures.print = 10)
```


```{r}
Elbow_plot <- ElbowPlot(HH30F, ndims = 40)
Elbow_plot
```


# find the neighbours 

```{r}
HH30F <- FindNeighbors(HH30F, dims = 1:30)

```

#find clusters
```{r}
HH30F <- FindClusters(HH30F, resolution = c(0.1, 0.2, 0.3, 0.4, 0.5, 0.6, 0.7, 0.8, 0.9))

```


#create the UMAP data - this the clusters the cells with common gene expression and links it to the neighbours

```{r}

HH30F <- RunUMAP(HH30F, dims = 1:30)
```
#trial

```{r}

res <- "RNA_snn_res."
nums = c(0.1, 0.2, 0.3, 0.4, 0.5)
paste0(res, nums)


DimPlot(HH30F, reduction = "umap", group.by = paste0(res, nums), label = TRUE, label.size = 6, repel = TRUE, pt.size = 0.5)
ggsave("/nobackup/proj/comet_icnlmdhuc/Taran/chicken_RNA_seq/HH30/HH30F/plots/cc/reg-full/cc_umap_trial.1.png", dpi = 400)



```

#visualise the UMAP

```{r}

res <- "RNA_snn_res."
nums = c(0.6, 0.7, 0.8, 0.9)
paste0(res, nums)


DimPlot(HH30F, reduction = "umap", group.by = paste0(res, nums), label = TRUE, label.size = 6, repel = TRUE, pt.size = 0.5)
ggsave("/nobackup/proj/comet_icnlmdhuc/Taran/chicken_RNA_seq/HH30/HH30F/plots/cc/reg-full/cc_umap_trial.2.png", dpi = 400)



```



```{r}

res <- "RNA_snn_res."
nums = c(0.4, 0.5, 0.7)
paste0(res, nums)


DimPlot(HH30F, reduction = "umap", group.by = paste0(res, nums), label = TRUE, label.size = 6, repel = TRUE, pt.size = 0.5)
ggsave("/nobackup/proj/comet_icnlmdhuc/Taran/chicken_RNA_seq/HH30/HH30F/plots/cc/reg-full/cc_umap_trial.3.png", dpi = 400)

```

#save as png
```{r}


DimPlot(HH30F, reduction = "umap", group.by = "RNA_snn_res.0.7", label = TRUE, label.size = 10, repel = TRUE, pt.size = 0.5) 
ggsave("/nobackup/proj/comet_icnlmdhuc/Taran/chicken_RNA_seq/HH30/HH30F/plots/cc/reg-full/cc_f.reg_umap-1.png", dpi = 400)

```



#update how clusters are ordered in metadata to ensure uniform ordering across all sample reports
```{r}
# Update the identity of the Seurat object and define the new order of groups
HH30F@meta.data$Phase <- factor(HH30F@meta.data$Phase, levels = c("G1", "S", "G2M"))


# Reassign identity with the new order
Idents(HH30F) <- HH30F@meta.data$Phase
```

#plot

```{r}

DimPlot(HH30F, cols = c("S" = "orange", "G1" = "forestgreen", "G2M" = "steelblue4"), group.by = "Phase", pt.size = 0.5) 
ggsave("/nobackup/proj/comet_icnlmdhuc/Taran/chicken_RNA_seq/HH30/HH30F/plots/cc/reg-full/cc_reg_full.png", dpi = 400)

```



```{r}
saveRDS(HH30F, file =  "/nobackup/proj/comet_icnlmdhuc/Taran/chicken_RNA_seq/HH30/HH30F/sObj/sObj_6_full_reg.RDS")
```
