---
title: "setup seurat object"
output: html_document
date: "2025-11-25"
fig_width: 6
fig_height: 4
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

#load all libaries required for this script

```{r} 
library(dplyr)
library(Seurat)
library(patchwork)
library(ggplot2)
library(DoubletFinder)
library(tidyverse)
library(RColorBrewer)
library(cowplot)
library(openxlsx)

# colour palette 
my_cols <- brewer.pal(n=5, "Spectral")


## feature and cell count function

FeatCellCount <- function(seuratObj){dim(seuratObj)
              }
  
  
##create function for doublet finder 


preDoublets <- function(seuratObj){  
    seuratObj <- NormalizeData(seuratObj)
    
    seuratObj <- ScaleData(seuratObj)
    seuratObj <- FindVariableFeatures(seuratObj, 
                                      x.low.cutoff = 0.0125, 
                                      y.cutoff = 0.25, 
                                      do.plot=FALSE)
    seuratObj <- RunPCA(seuratObj, pc.genes = seuratObj@var.genes, pcs.print = 0)
    
 
    return(seuratObj)
}



findDoublets <- function(seuratObj, n_pcs=30){
    
   
    nExp_poi  <- round(0.15*nrow(seuratObj@meta.data))
    
    print('yay!')
  
    seuratObj <- doubletFinder(seuratObj, 
                                  PCs = 1:30, 
                                  pN = 0.25, 
                                  pK = 0.01, 
                                  nExp = nExp_poi, 
                                  sct=FALSE)
    
    
    
     seuratObj@meta.data <- seuratObj@meta.data %>% 
        rename_at(vars(starts_with("DF.classifications")), 
                  funs(str_replace(., ".*", "DF.classifications"))) %>%
        rename_at(vars(starts_with("pANN")), 
                  funs(str_replace(., ".*", "pANN")))
    
       
  
        return(seuratObj) 
    
}


#create a workbook that contains assigns each clusters data to its own sheet

cluster_workbook <- function(data, n_clusters){
  
  wb <- createWorkbook()
  addWorksheet(wb, "Markers")
writeData(wb, sheet = "Markers", x = filtered_markers)

  for(i in 0:n_clusters){
    
    cluster_data <- filtered_markers[filtered_markers$cluster == i,] 
    addWorksheet(wb, as.character(i))
    writeData(wb, sheet = as.character(i), x = cluster_data)
    saveWorkbook(wb, file = "filtered_markers.xlsx", overwrite = TRUE)
    
  }
  
  
}



```

#load scRNA-seq data set 

```{r}

HH30F.data <- Read10X(data.dir = "/nobackup/proj/comet_icnlmdhuc/Taran/chicken_RNA_seq/results/1_HH30_F_results/outs/filtered_feature_bc_matrix/")
```

#assign to seurat object

```{r}

HH30F <- CreateSeuratObject(counts = HH30F.data, project = "HH30", min.cells = 3, min.features = 200)
HH30F


```


#add column to metadata to assign the percentage of mitochondrial dna to

```{r}
HH30F[["percent.mt"]] <- PercentageFeatureSet(HH30F, pattern = "^MT-")
```

#visualise QC metrics as violin plot

```{r}
QC_metric_violin_plot<- VlnPlot(HH30F, features = c("nFeature_RNA", "nCount_RNA", "percent.mt"), ncol = 3)

QC_metric_violin_plot

```

#feature scatter used to visualise the feature-feature relationships but can be used for any relationship

```{r}

mt_genes_scatter_plot1 <- FeatureScatter(HH30F, feature1 = "nCount_RNA", feature2 = "percent.mt")
features_scatter_plot2 <- FeatureScatter(HH30F, feature1 = "nCount_RNA", feature2 = "nFeature_RNA")

mt_genes_scatter_plot1 + features_scatter_plot2
```






#subset data to look at specific parameters bc values that are out of scope aren't important and therefore not needed in the graph (??? help)
#remove mitochondrial genes


```{r}
HH30F <- subset(HH30F, subset = nFeature_RNA > 250 & 
                                percent.mt   <5     &
                                nCount_RNA   >1000  
                )

mt_genes_scatter_plot3 <- FeatureScatter(HH30F, feature1 = "nCount_RNA", feature2 = "percent.mt")
features_scatter_plot4 <- FeatureScatter(HH30F, feature1 = "nCount_RNA", feature2 = "nFeature_RNA")

mt_genes_scatter_plot3 + features_scatter_plot4
```
#compare mitchondrial RNA subset data

```{r}
mt_genes_scatter_plot1 + mt_genes_scatter_plot3
```

#compare features subset data

```{r}
features_scatter_plot2 + features_scatter_plot4

```
#doubletFinder

```{r}

HH30F <- preDoublets(HH30F)
HH30F <- findDoublets(HH30F)

```


```{r}
HH30F <- subset(HH30F, subset = DF.classifications != "Doublet" )
```



#plot after doublet finder comparison

```{r}
mt_genes_scatter_plot5 <- FeatureScatter(HH30F, feature1 = "nCount_RNA", feature2 = "percent.mt")
features_scatter_plot6 <- FeatureScatter(HH30F, feature1 = "nCount_RNA", feature2 = "nFeature_RNA")

```

#compare mito RNA before and after subset and after doublet finder

```{r}

mt_genes_scatter_plot1 + mt_genes_scatter_plot3 + mt_genes_scatter_plot5

```

```{r}

features_scatter_plot2 + features_scatter_plot4 + features_scatter_plot6
```




#checkpoint after QC 

```{r}
saveRDS(HH30F, file = "sObj_QC_1.RDS")
```




#normalise the data

```{r}
HH30F <- NormalizeData(HH30F, normalization.method = "LogNormalize", scale.factor = 10000)
```

#find the variable features


```{r}
HH30F <- FindVariableFeatures(HH30F, selection.method = "vst", nfeatures = 2000)
```

#identify top 10 variable features

```{r}
top10 <- head(VariableFeatures(HH30F), 10)
```

#plot variable features

```{r}

variable_feat_plot <- VariableFeaturePlot(HH30F)
labelled_variable_feat_plot <- LabelPoints(plot = variable_feat_plot, points = top10, repel = TRUE, xnudge = 0, ynudge = 0)

```

#view plot

```{r}
variable_feat_plot +labelled_variable_feat_plot
```

# scale data so values are workable and are of same meaning

```{r}
all.genes <- rownames(HH30F)
HH30F <- ScaleData(HH30F, features = all.genes)
```


#run a principle component analysis to get our principle components

```{r}
HH30F <- RunPCA(HH30F, features = VariableFeatures(object = HH30F))
```

#look at first 5 PCAs

```{r}
print(HH30F[["pca"]], dims = 1:5, nFeatures = 5)
```

#do a heatmap to look at first PC

```{r}

DimHeatmap(HH30F, dims = 1, cells = 500, balanced = TRUE, fast = FALSE) + scale_fill_gradientn(colours = my_cols)



```
```{r}

PC1t5 <- DimHeatmap( HH30F, dims = 1:5,cells = 500, balanced  = TRUE, fast = FALSE, combine = FALSE)

PC1t5_col <- lapply(PC1t5, function(p) p + scale_fill_gradientn(colours = my_cols))

wrap_plots(plotlist = PC1t5_col, nrow = 3, ncol = 2)


```

#checkpoint after PCA

```{r}

saveRDS(HH30F, file = "sObj_PCA_2.RDS")

```


#elbow plot to determine dimensionality for the cluster map

```{r}
Elbow_plot <- ElbowPlot(HH30F, ndims = 40)
Elbow_plot
```



# cluster cells and find the neighbours of the clusters

```{r}
HH30F <- FindNeighbors(HH30F, dims = 1:30)
HH30F <- FindClusters(HH30F, resolution = 0.8)

```

#create the UMAP data - this the clusters the cells with common gene expression and links it to the neighbours

```{r}

HH30F <- RunUMAP(HH30F, dims = 1:30)
```

#visualise the UMAP

```{r}


umap_HH30 <- DimPlot(HH30F, reduction = "umap", label = TRUE, repel = TRUE)

umap_HH30
```

#save RDS

```{r}
saveRDS(HH30F, file = "sObj_UMAP_3.rds")
```



# find markers for every cluster compared to all remaining cells, report only the positive ones, write to excel file
```{r}
markers <- FindAllMarkers(HH30F, only.pos = TRUE)
markers %>%
    group_by(cluster) %>%
    dplyr::filter(avg_log2FC > 1)


write.xlsx(markers, file = "HH30F_markers.xlsx")

```


#filter out "ENSGAL" genes

```{r}

filtered_markers <- markers %>% filter(!startsWith(gene, "ENSGAL"))

```



#write the markers to a workbook

```{r}

cluster_workbook(filtered_markers, 16)

```



#several tests for differential expression - set parameter test.use = "ROC" bc it will return the classification power for any individual marker
#find differential expression of markers for cluster 0 and write to excel file

```{r}
cluster0.markers <- FindMarkers(HH30F, ident.1 = 0, logfc.threshold = 0.25, test.use = "roc", only.pos = TRUE)

write.xlsx(cluster0.markers, file = "HH30F_c0_markers.xlsx")
```


```{r}

saveRDS(HH30F, file = "sObj_Markers_4.RDS")

```

