---
title: "cell-cycle-regression"
author: "Taran"
date: "2026-07-07"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```




```{r} 
library(dplyr)
library(Seurat)
library(patchwork)
library(ggplot2)
library(DoubletFinder)
library(tidyverse)
library(RColorBrewer)
library(cowplot)
library(openxlsx)

# colour palette 
my_cols <- brewer.pal(n=5, "Spectral")


## feature and cell count function

FeatCellCount <- function(seuratObj){dim(seuratObj)
              }


```


#find difference between scores
```{r}
HH28C$CC.Difference <- HH28C$S.Score - HH28C$G2M.Score
```

#regress the difference
```{r}
HH28C <- ScaleData(HH28C, vars.to.regress = "CC.Difference", features = VariableFeatures(HH28C))
```

# cell cycle effects removed in PCA
```{r}
HH28C <- RunPCA(HH28C, features = VariableFeatures(HH28C), nfeatures.print = 10)
```



```{r}
Elbow_plot <- ElbowPlot(HH28C, ndims = 40)
Elbow_plot
```


# find the neighbours 

```{r}
HH28C <- FindNeighbors(HH28C, dims = 1:30)

```

#find clusters
```{r}
HH28C <- FindClusters(HH28C, resolution = c(0.1, 0.2, 0.3, 0.4, 0.5, 0.6, 0.7, 0.8, 0.9))
```

#create the UMAP data - this clusters the cells with common gene expression and links it to the neighbours

```{r}

HH28C <- RunUMAP(HH28C, dims = 1:30)
```

#trial

```{r}

res <- "RNA_snn_res."
nums = c(0.1, 0.2, 0.3, 0.4, 0.5)
paste0(res, nums)


DimPlot(HH28C, reduction = "umap", group.by = paste0(res, nums), label = TRUE, label.size = 6, repel = TRUE, pt.size = 0.5)
ggsave("/nobackup/proj/comet_icnlmdhuc/Taran/chicken_RNA_seq/HH28/sample_3/plots/cc/reg-diff/cc_umap_trial.1.png", dpi = 400)



```

#visualise the UMAP

```{r}

res <- "RNA_snn_res."
nums = c(0.6, 0.7, 0.8, 0.9)
paste0(res, nums)


DimPlot(HH28C, reduction = "umap", group.by = paste0(res, nums), label = TRUE, label.size = 6, repel = TRUE, pt.size = 0.5)
ggsave("/nobackup/proj/comet_icnlmdhuc/Taran/chicken_RNA_seq/HH28/sample_3/plots/cc/reg-diff/cc_umap_trial.2.png", dpi = 400)



```



```{r}

res <- "RNA_snn_res."
nums = c(0.5, 0.6)
paste0(res, nums)


DimPlot(HH28C, reduction = "umap", group.by = paste0(res, nums), label = TRUE, label.size = 6, repel = TRUE, pt.size = 0.5)
ggsave("/nobackup/proj/comet_icnlmdhuc/Taran/chicken_RNA_seq/HH28/sample_3/plots/cc/reg-diff/cc_umap_trial.3.png", dpi = 400)

```

#save as png
```{r}


DimPlot(HH28C, reduction = "umap", group.by = "RNA_snn_res.0.5", label = TRUE, label.size = 5, repel = TRUE, pt.size = 0.5) 
ggsave("/nobackup/proj/comet_icnlmdhuc/Taran/chicken_RNA_seq/HH28/sample_3/plots/cc/reg-diff/cc_d.reg_umap1.png", dpi = 400)

```

#update how clusters are ordered in metadata to ensure uniform ordering across all sample reports
```{r}
# Update the identity of the Seurat object and define the new order of groups
HH28C@meta.data$Phase <- factor(HH28C@meta.data$Phase, levels = c("G1", "S", "G2M"))


# Reassign identity with the new order
Idents(HH28C) <- HH28C@meta.data$Phase
```


#plot

```{r}

DimPlot(HH28C, cols = c("S" = "orange", "G1" = "forestgreen", "G2M" = "steelblue4"), group.by = "Phase", pt.size = 0.5) 
ggsave("/nobackup/proj/comet_icnlmdhuc/Taran/chicken_RNA_seq/HH28/sample_3/plots/cc/reg-diff/cc_reg_diff.png", dpi = 400)


```


```{r}
saveRDS(HH28C, file = "/nobackup/proj/comet_icnlmdhuc/Taran/chicken_RNA_seq/HH28/sample_3/sObj/sObj_7_reg_diff.RDS")
```

