---
title: "markers"
author: "Taran"
date: "2026-07-07"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


```{r} 
library(dplyr)
library(Seurat)
library(patchwork)
library(ggplot2)
library(DoubletFinder)
library(tidyverse)
library(RColorBrewer)
library(cowplot)
library(openxlsx)

# colour palette 
my_cols <- brewer.pal(n=5, "Spectral")


## feature and cell count function

FeatCellCount <- function(seuratObj){dim(seuratObj)
              }


#create a workbook that contains assigns each clusters data to its own sheet

cluster_workbook <- function(data, n_clusters){
  
  wb <- createWorkbook()
  addWorksheet(wb, "Markers")
writeData(wb, sheet = "Markers", x = filtered_markers)

  for(i in 0:n_clusters){
    
    cluster_data <- filtered_markers[filtered_markers$cluster == i,] 
    addWorksheet(wb, as.character(i))
    writeData(wb, sheet = as.character(i), x = cluster_data)
    saveWorkbook(wb, file = "filtered_markers.xlsx", overwrite = TRUE)
    
  }
  
  
}


#top 5 genes for each cluster, sorted largest to smallest for avg log, more than 0.1 for pct.1 and less than 0.05 for p val adj
top5 <- function(sObj){

clusters <- filtered_markers %>% group_by(cluster) %>% arrange(cluster, desc(avg_log2FC)) %>% filter(pct.1 > 0.1) %>% filter(p_val_adj < 0.05) %>% top_n(5, avg_log2FC)
  

return(clusters)  
 
   
}


```

# find markers for every cluster compared to all remaining cells, report only the positive ones, write to excel file
```{r}
markers <- FindAllMarkers(HH28D, only.pos = TRUE, group.by = "RNA_snn_res.0.6")
markers %>%
    group_by(cluster) %>%
    dplyr::filter(avg_log2FC > 1)


write.xlsx(markers, file = "HH28D_markers.xlsx")

```


#filter out "ENSGAL" genes

```{r}

filtered_markers <- markers %>% filter(!startsWith(gene, "ENSGAL"))

```



#write the markers to a workbook

```{r}

cluster_workbook(filtered_markers, 13)

```



#several tests for differential expression - set parameter test.use = "ROC" bc it will return the classification power for any individual marker
#find differential expression of markers for cluster 0 and write to excel file

```{r}
cluster0.markers <- FindMarkers(HH28D, ident.1 = 0, logfc.threshold = 0.25, test.use = "roc", only.pos = TRUE)

write.xlsx(cluster0.markers, file = "HH28D_c0_markers.xlsx")
```



#look for marker expression throughout UMAP plot to identify clusters

```{r}

Progenitor_markers <- FeaturePlot(HH28D, features = c("ATOH7", "HES6"))

amacrine <- FeaturePlot(HH28D, features = c("TFAP2A", "TFAP2B", "PROX1", "LHX1"))

early_cone_photoreceptors <- FeaturePlot(HH28D, features = c("RAX2", "RXRG", "CRX"))

neural_retina <- FeaturePlot(HH28D, features = c("SMOC1", "ALDH1A3", "VAX1", "VSX2", "CDH2"))

related_to_RPE_and_RPC <- FeaturePlot(HH28D, features = c("TYR", "TYRP", "WNT2B", "BCN"))

retinal_ganglion_cells <- FeaturePlot(HH28D, features = c("ELAVL4", "NEFL", "NEFM", "CL12", "GAP43", "STMN", "POU4F3", "SATB1", "SATB2", "PAX6", "RBPMS2"))

retinal_progenitor_cells <- FeaturePlot(HH28D, features = c("ZIC1", "VSX2", "PAX6", "VAX1", "SMOC1", "PMPR1B", "ATOH7", "HES6"))

endothelial_and_blood <- FeaturePlot(HH28D, features = c("VWF", "KDR"))

POM <- FeaturePlot(HH28D, features = c("PITX2", "EMX2", "LMX1B", "PAX7", "ALX1", "LMX1B", "ALX4", "DCN", "LTBP1", "COL3A1"))

RPE <- FeaturePlot(HH28D, features = c("OTX2", "MITF", "TYR", "TYRP1", "PITX2", "PMEL", "CDH1"))

pioneer <- FeaturePlot(HH28D, features = c("PAX2", "CLYBL", "NTN1", "MET", "EGF", "DVL1", "GALNT6", "FOXP2", "CDH2"))

blood <- FeaturePlot(HH28D, features = c("HBE1", "HBBA", "HBAD", "HBA1", "HBE", "HBZ", "HBBR", "SLC4A1", "SLC39A8", "TSPO2"))


cmz <- FeaturePlot(HH28D, features = c("COL9A1", "CPAMD8", "MECOM", "OPTC", "RELN", "BNC", "FBN2", "FBN3"))




```

```{r}

Progenitor_markers 

amacrine 

early_cone_photoreceptors 

neural_retina 

related_to_RPE_and_RPC 

retinal_ganglion_cells 

retinal_progenitor_cells 

endothelial_and_blood 

POM 

RPE 

pioneer 

blood


cmz 


```

#top 5 genes for each cluster, sorted largest to smalled avg log and less than 0.05 for p val adj
```{r}
top5.genes <- top5(HH28D)

top5.genes
``` 


#heatmap for the top 5 genes per cluster

```{r}
top5heatmap <- DoHeatmap(HH28D, group.by = "RNA_snn_res.0.6", features = top5.genes$gene)
top5heatmap

```


```{r}

options(bitmapType = "cairo")
png("top5genes.png", width = 9000, height = 4000, res = 200, type = "cairo") 
p <- DoHeatmap(HH28D, group.by = "RNA_snn_res.0.7", features = top5.genes$gene, label = FALSE) 
p <- p + theme(
  axis.text = element_text(size = 20),         #gene names
  legend.title = element_text(size = 20),      #identity and expression text size
  legend.text = element_text(size = 20),       #key label size
  legend.key.size = unit(1.5, "cm"))           #box size
  
wrap_plots(p)

dev.off()
```



#assign cluster identity


```{r}
new.cluster.ids <- c("0", "1", "2", "RPE", "CMZ", "5", "RPE2", "Pioneer", "8","POM","RPC", "EC", "12", "13" )


Idents(HH28D) <- HH28D@meta.data$RNA_snn_res.0.6

names(new.cluster.ids) <- levels(HH28D)

HH28D <- RenameIdents(HH28D, new.cluster.ids)
DimPlot(HH28D, reduction = "umap", label = TRUE, label.size = 8) + xlab("UMAP 1") + ylab("UMAP 2") +
    theme(axis.title = element_text(size = 18), legend.text = element_text(size = 18)) + guides(colour = guide_legend(override.aes = list(size = 10)))
```




```{r}

png(file= "umap_HH28D.png", width=2800, height=1500, res = 200, type = "cairo") 
DimPlot(HH28D, reduction = "umap", label = TRUE, label.size = 8) + xlab("UMAP 1") + ylab("UMAP 2") +
    theme(axis.title = element_text(size = 18), legend.text = element_text(size = 18)) + guides(colour = guide_legend(override.aes = list(size = 10)))
dev.off()


```



```{r}

saveRDS(HH28D, file = "sObj_Markers_4.RDS")

```
