---
title: "reg-full"
author: "Taran"
date: "2026-08-04"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```



```{r} 
library(dplyr)
library(Seurat)
library(patchwork)
library(ggplot2)
library(DoubletFinder)
library(tidyverse)
library(RColorBrewer)
library(cowplot)
library(openxlsx)

# colour palette 
my_cols <- brewer.pal(n=5, "Spectral")


## feature and cell count function

FeatCellCount <- function(seuratObj){dim(seuratObj)
              }


```


#segregate list of cell cycle markers into G2/M and S

```{r}
s.genes <- cc.genes$s.genes
g2m.genes <- cc.genes$g2m.genes
```



```{r}

HH30G <- ScaleData(HH30G, vars.to.regress = c("S.Score", "G2M.Score"), features = VariableFeatures(HH30G)) 

```


#PCA on the variable genes no longer returns components associated with cell cycle

```{r}
HH30G <- RunPCA(HH30G, features = VariableFeatures(HH30G), nfeatures.print = 10)
```


```{r}
Elbow_plot <- ElbowPlot(HH30G, ndims = 40)
Elbow_plot
```


# find the neighbours 

```{r}
HH30G <- FindNeighbors(HH30G, dims = 1:30)

```

#find clusters
```{r}
HH30G <- FindClusters(HH30G, resolution = c(0.4, 0.5, 0.6, 0.7))

```


#create the UMAP data - this the clusters the cells with common gene expression and links it to the neighbours

```{r}

HH30G <- RunUMAP(HH30G, dims = 1:30)
```
#trial

```{r}

res <- "RNA_snn_res."
nums = c(0.4, 0.5, 0.6, 0.7)
paste0(res, nums)


p <- DimPlot(HH30G, reduction = "umap", group.by = paste0(res, nums), label = TRUE, label.size = 6, repel = TRUE, pt.size = 0.5)
LabelClusters(p, id = "RNA_snn_res.0.6", repel = TRUE)


ggsave("/nobackup/proj/comet_icnlmdhuc/Taran/chicken_RNA_seq/HH30/HH30G/plots/cc/reg-full/cc_umap_trial.1.png", dpi = 400, width =45, height = 25, units = "cm")



```




```{r}

res <- "RNA_snn_res."
nums = c(0.5, 0.6)
paste0(res, nums)


DimPlot(HH30G, reduction = "umap", group.by = paste0(res, nums), label = TRUE, label.size = 6, repel = TRUE, pt.size = 0.5)
ggsave("/nobackup/proj/comet_icnlmdhuc/Taran/chicken_RNA_seq/HH30/HH30G/plots/cc/reg-full/cc_umap_trial.3.png", dpi = 400, width =45, height = 25, units = "cm")

```

#save as png
```{r}


DimPlot(HH30G, reduction = "umap", group.by = "RNA_snn_res.0.6", label = TRUE, label.size = 10, repel = TRUE, pt.size = 0.5) 
ggsave("/nobackup/proj/comet_icnlmdhuc/Taran/chicken_RNA_seq/HH30/HH30G/plots/cc/reg-full/cc_f.reg_umap-1.png", dpi = 400, width =45, height = 25, units = "cm")

```



#update how clusters are ordered in metadata to ensure uniform ordering across all sample reports
```{r}
# Update the identity of the Seurat object and define the new order of groups
HH30G@meta.data$Phase <- factor(HH30G@meta.data$Phase, levels = c("G1", "S", "G2M"))


# Reassign identity with the new order
Idents(HH30G) <- HH30G@meta.data$Phase
```

#plot

```{r}

DimPlot(HH30G, cols = c("S" = "orange", "G1" = "forestgreen", "G2M" = "steelblue4"), group.by = "Phase", pt.size = 0.5) 
ggsave("/nobackup/proj/comet_icnlmdhuc/Taran/chicken_RNA_seq/HH30/HH30G/plots/cc/reg-full/cc_reg_full.png", dpi = 400, width =45, height = 25, units = "cm")

```



```{r}
saveRDS(HH30G, file =  "/nobackup/proj/comet_icnlmdhuc/Taran/chicken_RNA_seq/HH30/HH30G/sObj/sObj_6_full_reg.RDS")
```
